Relationship between solar wind, $D_{st}$, and plasmasphere mass density on one-hour time scales

### 2015-06-18 paper.tex
Added significance tests and NaN tests

\twocolumn[
  \begin{@twocolumnfalse}
\maketitle
\hrule
\begin{abstract}
This paper looks at how various magnetosphere conditions behave around the onset of geomagnetic storms. It confirms results from previous papers that a sudden drop in $D_{st}$ correlates with a spike in equatorial mass density, while adding a level of depth and specification to the applicability of their results. It compares data at 1-hour averages to data at 1-day averages, and briefly to data at 27-day averages to examine if the trend holds at varying time scales and finds that it does, so long as the data is pre-processed to linearly remove a large time scale $F_{10.7}$ dependence. 
By then looking at an hourly average of variables from 24 hours before storm onset to 48 hours after, short timescale trends can be discerned.

@@

In this work events are considered when $D_{st}$ or $\rho_{eq}$ crosses a threshold value.  In finding events, all fill values were replaced with linearly interpolated values.  
% unless the gap was larger than X hours?

@@

% http://www.dartmouth.edu/~rdenton/Data/DentonTakahashiGOES1980-1991MassDensityWithHeader.txt
% This looks like a 10 minute grid, not 15 minute?

@@

\subsection{D$_{st}$ Events}
Two storm indicators are looked at in this study. The first is looking for a drop in DST below the threshhold of $-50nT$ specified in \cite{Takahashi2010}, dubbed the "onset", and then considering the timeframe a storm until $D_{st}$ passes back above the $-50nT$ threshold. This method finds 669 such periods between May 1983 and August 1991 with an average duration of 9 hours and a median duration of 3 hours. A subset of longer duration storms will be looked at later. Figure \ref{AllData} shows the average values of $B_Z$, $V_{SW}$, $D_{st}$, $F_{10.7}$, and Mass Density for a period of 24 hours before and 48 hours after a storm onsppet, as well as a plot of mass density with the found storm times highlighted.

@@

\subsection{Mass Density Events}
Figure \ref{MassStorm} shows this same algorithm, but looking for a rise in mass density over a value of 40$g/cm^3$. This results in 130 storms with a mean duration of 32 hours and a median duration of 17 hours, marked in red on the left figure.

This shows that when using all data for mass density derived storms, almost no significant changes can be seen around storm onset.

\section{More specific storms}
It's hypothesized that progressively picking more specific storm criteria will allow for the possibility of more significant results, at the expense of more bias in the selection process and potentially less overall usefulness of the results. That said, the predictability of extreme storms is of definite interest, so an attempt has been made to find some reproducible method of prediction. Looking at storms that last longer than 12 hours and storms with an onset threshhold greater than $70g/cm^3$ results in the left and right sides of Figure \ref{Mspec} respectively.

Neither of these seem to indicate anything too significant, so looking at $D_{st}$ storms instead to look for something that causes a significant change in Mass Density results in Figure \ref{Dspec}. This shows that by either looking only at $D_{st}$ storms that last longer than an hour (left) or at storms where the onset condition is $D_{st}<-80nT$, a spike in mass density is seen, but also a definite lack of data availability to the point where that spike may be coming from less than five of the total 143 storms.

Unforunately there are no storms in this time frame that are longer than 12 hours with $D_{st}$ minima lower than $-80nT$ that have existing mass density data around onset, so an analysis of this particular relationship can't be made.

\subsection{Change in $\rho_{eq}$}
If instead of looking for storms based on a threshold of $\rho_{eq}$, we instead look for a certain amount of change in $\rho_{eq}$ as the basis for storm onset, we get Figure \ref{rhochange}.

@@

Taking this re-created data set and subtracting it from the original should remove the $F_{10.7}$ dependence from the data, and allow for a less biased analysis of the relationship between $D_{st}$ and $\rho_{eq}$. Figure \ref{nof107} shows the stack plot for the reduced $F_{10.7}$, where a more distinct peak after storm onset can be seen. It also shows what that removed trend looks like, of the form $\rho_{eq}-\rho_{eq,F_{10.7}}$.

@@

@@

% Remove title from both figures.  D_ST -> D_st Bz -> B_z in labels.  Change non-nan data points to ``# of values''.
% Change x-label to be ``Time from start of event (hr)'' and put vertical lines in each panel at t = 0 (somewhere I have a code called vertline.m)
\begin{figure}[htp]
\centering
(a)
\includegraphics[scale=0.7]{paperfigures/massdensitystorms.eps}
(b)
\includegraphics[scale=0.7]{paperfigures/dststorms.eps}
\caption{Averages around time of (a) 669 $\rho_{eq}$ and (b) 130 $D_{st}$ events along with thresholds used to define an event. The number of values used in the averages for mass density is shown in the bottom panel (a) and (b).}
\label{DSTstorm}
\end{figure}
\clearpage

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.7]{paperfigures/stormavs-md12.png}
\includegraphics[scale=0.7]{paperfigures/stormavs-m70.png}
\caption{(a) Same as \ref{fig:MassStorm}(a) except with constraint that D$_{st}$ stayed below -40~nT for at least 12~hours. (b) Same as \ref{fig:MassStorm}(b) except for constrain that mass density crossed above $70 amu/cm^3$.}
\label{Mspec}
\end{figure}
\clearpage

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.7]{paperfigures/stormavs-dd12.png}
\includegraphics[scale=0.7]{paperfigures/stormavs-d80.png}
\caption{$D_{st}$ Storms $>$ 12 hours, $D_{st}<-80nT$}
\label{Dspec}
\end{figure}
\clearpage

>>>>

\twocolumn[
  \begin{@twocolumnfalse}
\maketitle
\hrule
\begin{abstract}
This paper looks at how various magnetosphere conditions behave around the onset of geomagnetic events. It confirms results from previous papers that a sudden drop in $D_{st}$ correlates with a spike in equatorial mass density, while adding a level of depth and specification to the applicability of their results. It compares data at 1-hour averages to data at 1-day averages, and briefly to data at 27-day averages to examine if the trend holds at varying time scales and finds that it does, so long as the data is pre-processed to linearly remove a large time scale $F_{10.7}$ dependence. 
By then looking at an hourly average of variables from 24 hours before event onset to 48 hours after, short timescale trends can be discerned.

@@

In this work events are considered when $D_{st}$ or $\rho_{eq}$ crosses a threshold value.  In finding events, all fill values were replaced with linearly interpolated values.

@@

To check for potential bias in the data as used by our analysis, we checked how the data availability varied with hour, shown in Figure \ref{nanperhour}. We also looked at how storm conditions affected data availability. Both a two sample t-test for difference in means and a Wilcoxon rank sum test showed significance at the 1\% level that $D_{st}$ during periods with available data was of a different distribution than $D_{st}$ during periods missing data. When testing for other significant $D_{st}$ differences, the Wilcoxon test was significant while the t-test was not in pre-noon vs post-noon distributions, as well as $\rho_{eq}>40$ vs $\rho_{eq}<40$ distributions indicating perhaps an increase in variability without an increase in mean.

@@

\subsection{D$_{st}$ Events}
Two event indicators are looked at in this study. The first is looking for a drop in DST below the threshhold of $-50nT$ specified in \cite{Takahashi2010}, dubbed the "onset", and then considering the timeframe a event until $D_{st}$ passes back above the $-50nT$ threshold. This method finds 669 such periods between May 1983 and August 1991 with an average duration of 9 hours and a median duration of 3 hours. A subset of longer duration events will be looked at later. Figure \ref{AllData} shows the average values of $B_Z$, $V_{SW}$, $D_{st}$, $F_{10.7}$, and Mass Density for a period of 24 hours before and 48 hours after a event onsppet, as well as a plot of mass density with the found event times highlighted.

@@

\subsection{Mass Density Events}
Figure \ref{MassStorm} shows this same algorithm, but looking for a rise in mass density over a value of 40$g/cm^3$. This results in 130 events with a mean duration of 32 hours and a median duration of 17 hours, marked in red on the left figure.

This shows that when using all data for mass density derived events, almost no significant changes can be seen around event onset.

\section{More specific events}
It's hypothesized that progressively picking more specific event criteria will allow for the possibility of more significant results, at the expense of more bias in the selection process and potentially less overall usefulness of the results. That said, the predictability of extreme events is of definite interest, so an attempt has been made to find some reproducible method of prediction. Looking at events that last longer than 12 hours and events with an onset threshhold greater than $70g/cm^3$ results in the left and right sides of Figure \ref{Mspec} respectively.

Neither of these seem to indicate anything too significant, so looking at $D_{st}$ events instead to look for something that causes a significant change in Mass Density results in Figure \ref{Dspec}. This shows that by either looking only at $D_{st}$ events that last longer than an hour (left) or at events where the onset condition is $D_{st}<-80nT$, a spike in mass density is seen, but also a definite lack of data availability to the point where that spike may be coming from less than five of the total 143 events.

Unforunately there are no events in this time frame that are longer than 12 hours with $D_{st}$ minima lower than $-80nT$ that have existing mass density data around onset, so an analysis of this particular relationship can't be made.

\subsection{Change in $\rho_{eq}$}
If instead of looking for events based on a threshold of $\rho_{eq}$, we instead look for a certain amount of change in $\rho_{eq}$ as the basis for event onset, we get Figure \ref{rhochange}.

@@

Taking this re-created data set and subtracting it from the original should remove the $F_{10.7}$ dependence from the data, and allow for a less biased analysis of the relationship between $D_{st}$ and $\rho_{eq}$. Figure \ref{nof107} shows the stack plot for the reduced $F_{10.7}$, where a more distinct peak after event onset can be seen. It also shows what that removed trend looks like, of the form $\rho_{eq}-\rho_{eq,F_{10.7}}$.

@@

\begin{figure}[htp]
\centering
\includegraphics[scale=0.5]{paperfigures/nansbyhour.eps}
\includegraphics[scale=0.5]{paperfigures/nansbyhour_storm.eps}
\caption{(a) Number of NaN points per hour of observation in the total data set (b) Data availability relative to average event hour in events where $D_{st}<-80nT$}
\label{nanperhour}
\end{figure}
\clearpage

@@

% Change x-label to be ``Time from start of event (hr)'' and put vertical lines in each panel at t = 0 (somewhere I have a code called vertline.m)
\begin{figure}[htp]
\centering
(a)
\includegraphics[scale=0.7]{paperfigures/massdensitystorms.eps}
(b)
\includegraphics[scale=0.7]{paperfigures/dststorms.eps}
\caption{Averages around time of (a) 669 $\rho_{eq}$ and (b) 130 $D_{st}$ events along with thresholds used to define an event. The number of values used in the averages for mass density is shown in the bottom panel (a) and (b).}
\label{DSTstorm}
\end{figure}
\clearpage

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.5]{paperfigures/stormavs-dd12.eps}
\includegraphics[scale=0.5]{paperfigures/stormavs-d80.eps}
\caption{(a) Same as \ref{fig:MassStorm}(a) except with constraint that $D_{st}$ stayed below -40~nT for at least 12~hours. (b) Same as \ref{fig:MassStorm}(b) except for constrain that $D_{st}$ crossed below $-80 nT$.}
\label{Mspec}
\end{figure}
\clearpage

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.5]{paperfigures/stormavs-md12.eps}
\includegraphics[scale=0.5]{paperfigures/stormavs-m70.eps}
\caption{(a) Same as \ref{fig:MassStorm}(a) except with constraint that $\rho_{eq}$ stayed above 50~nT for at least 12~hours. (b) Same as \ref{fig:MassStorm}(b) except for constrain that mass density crossed above $70 amu/cm^3$.}
\label{Dspec}
\end{figure}
\clearpage
